### 2016-08-24 thesis_paper/source.tex
Some paper modifications
Draft version of thesis paper

\titleformat{\chapter}[display]{\normalfont\Large\bfseries\centering}{\chaptertitlename\ \thechapter}{0pt}{\Large\uppercase}

@@

% Make chapter titles uppercase is a pain.
% The internal chapter macros must be redefined.
\makeatletter
\renewcommand\chapter{\if@openright\cleardoublepage\else\clearpage\fi
                    \thispagestyle{plain}%
                    \global\@topnum\z@
                    \@afterindentfalse
                    \secdef\@chapter\@schapter}
\def\@chapter[#1]#2{\ifnum \c@secnumdepth >\m@ne
                       \if@mainmatter
                         \refstepcounter{chapter}%
                         \typeout{\@chapapp\space\thechapter.}%
                         \addcontentsline{toc}{chapter}%
                         {\protect\numberline{\thechapter}\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                       \else
                       \addcontentsline{toc}{chapter}{\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                       \fi
                    \else
                    \addcontentsline{toc}{chapter}{\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                    \fi
                    \chaptermark{#1}%
                    \addtocontents{lof}{\protect\addvspace{10\p@}}%
                    \addtocontents{lot}{\protect\addvspace{10\p@}}%
                    \if@twocolumn
                      \@topnewpage[\@makechapterhead{#2}]%
                    \else
                      \@makechapterhead{#2}%
                      \@afterheading
                    \fi}
\makeatother

\begin{document}
{
\thispagestyle{empty}

@@

\addcontentsline{toc}{chapter}{\abstractname}

@@

\begin{abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

@@

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier (it will be explained more in later sections). In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much, though it didn't bear any loss in precision and performance.
\end{abstract}

@@

\chapter{Introduction}

\justify
In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

\vspace{5mm}

@@

\vspace{3mm}
The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

@@

The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms respectively on Figure \ref{fg:collagen_dipal}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems with 890,000 atoms with 402,400 atoms}
 \label{fg:collagen_dipal}
\end{figure}

\section{Problem Statement}
Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

Reviewing the approach we are coming up with and what it was before helps us analyzing the weak parts and things that we can focus on to eliminate old bottlenecks.

@@

Finally, we've analyzed some main ideas, advantages, and disadvantages of our approach in this paper which is basically push-based system with optimization on data processing part by means of GPU computation (in other words, parallelism). Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. You can observe some estimation samples in Figure \ref{fg:sample_estimation_old_paper}. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

@@

In collaboration with Physics Department, the original motivation came from the need of improving analytic query executions performance over big data being generated by simulation software. The idea was to take input stream coming from simulation software (ideally live) and run network of queries very efficiently, because sometimes they might take days and more to compute. The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis fast having an ability to expand network of queries and customize computation functions to get the desired results. 
Our system primarily represents a system that works in several modules that were developed in C, CUDA, Python programming languages that, in combination, give us the desired result to perform efficient computations for the network of queries developed in the previous paper \cite{mainPaper} as well as to prove the concept of optimization of push-based system by means of parallelism with no data loss. We've developed the following modules:

@@

Push-based System for molecular simulation data analysis with its improved GPU features has couple of main focuses in terms of technical implementation and challenges related to those that have already been tried to be solved by other different approaches and software. Here is the review of other works.

@@

First of all, this system is a tool for scientific computation and in this case particularly for molecular simulation data analysis. PeriScope is a project and initiative by University of Michigan which is a system that will optimize data management for scientific explorations. Relational Database Management Systems were created to optimize business applications and it was revolutionary. Those couldn't be built without these systems with abstract languages like SQL, since it manipulates and automatically takes use of indexes, etc. PeriScope is a project that tries to come up with same approach against scientific data and computation with respectively larger scale.\cite{PeriScope} It is kind of difficult to compare our proposed system with PeriScope, but the original motivation is very similar: optimize scientific computation. PeriScope is a larger project directed towards more abstract solution like RDBMs that would work for any type of data. Another existing tool for scientific data and optimized processing is SciDB. It was created to make scientific data storage and querying better and faster. It is very similar to traditional RDBMS systems in terms of querying language and represent data storage in tables, but states to be faster than Postgres in 2 orders of magnitude. Some of the main features is distributed processing with atomicity and durability (the latter is quite difficult for large scale data), while the disadvantages are no-overwriting (only append mode), single value columns only (no arrays, etc), where records are like simple tuples with join function on querying multiple tables \cite{SciDB}. One of the major differences between Push-based Molecular simulation data analysis tool and the projects above is that our proposed design doesn't do anything with storing the data, because we are mainly concentrating it as a side plugin for processing and querying data in an optimized way, while the actual data generation and storage will be done with the actual simulation software.

\section{Stream processing}
Another major modern issue is live data streaming and live processing, which is also possible with our system. Basically, streaming function is a big challenge and need nowadays, because such networks as Twitter and Facebook are able to keep track of live trends and events, including people's reaction and feedback that have to be reflected real time for tremendous amount of users across the globe. Borealis Stream Processing Software is a tool being developed by MIT, Brown, and Brandeis Universities. It's primary goal is a distributed stream processing and based on Medusa and Aurora. Some of the major functions of the software are highly available distributed large streams processing live, dynamic and highly-scalable with possibility of changing queries live and dynamic vision of query results. \cite{borealis} There is also a similar need in scientific world, since being able to keep track of results in real time might be crucial to be able to make certain explorations.

@@

We are potentially dealing with data that could easily get up to terabytes and petabytes in sizes, and it will be fair to apply term of big data. Traditional software tools are not able to process Big Data apparently because of huge data sets, thus, special frameworks were developed to handle them. Nowadays there are bunch of frameworks that try to solve this problem. Some of these frameworks are Hadoop \cite{hadoop}, Hadoop with GPU extension \cite{hadoopgpu}, Spark \cite{spark}, Flink \cite{flink} (this is more of a big data stream processing framework) and other ones that are possibly just custom modifications and distributions of these. The mentioned tools help dealing with big data giving an ability to process them in batches on clusters using underlying HDFS (Highly Distributed File System), and normally go along with other utilities that help to apply MapReduce concept. The relationship between our system and this concept along with its implementations is in data sets sizes. Even though we are not using or introducing any non-traditional file systems, in certain scale and problems it could be required, too. In fact, the frameworks listed above could even be used for our problem on top of our system, because, as we discuss in section \ref{sc:conclusion}, it could be extended to a distributed system for analyzing independent frames in parallel.

\section{GPU and Big Data Processing}
GPU is becoming more popular being used by not only graphics processing, but machine learning, deep learning, search algorithms, sorting, and so on. Having an ability to optimize in tremendous speedups it is being involved in video classifications, natural language processing. Particularly about distributed processing and GPU applications along with it, there has been made some overview and research on what is the current state for common solutions in industry. "There were a lot of custom GPU-Hadoop frameworks after the launch of Mars project. These include C-MR, SteamMR, GPMR, Grex, Panda, Shredder and many others. However, most of these frameworks are no longer supported and were developed for some particular scientific projects. Therefore, a Monte Carlo simulation framework can be hardly applied for a bioinformatics project (say) based on the other algorithms. Moreover, processor technologies are evolving very rapidly. Many new revolutionary architectures have been developed for Sony PlayStation 4, Mali GPU by ARM, Adapteva Multicore Microprocessor etc. Both Mali GPU and Adapteva will be compatible with OpenCL. Xeon Phi co-processor works with OpenCL too, which was launched by INTEL. It has an x86-like architecture and is a 60 core co-processor that supports the PCI Express standard. It consumes just 300watts of energy and provides by giving a performance of 1 TFLOPS in double precision. Tianhe-2 which is the most powerful supercomputer till date implements this co-processor. Though, it is very difficult determine which Processor Computing framework architecture will more robust and high
performance." \cite{hadoopgpu}

@@

\chapter{Benchmark scientific queries}\label{sc:querynetwork}
Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries that were already introduced in the previous work of Push-Based System for Molecular Simulation Data Analysis \cite{mainPaper}.

@@

\begin{table}[h]
\centering
\resizebox{\columnwidth}{!}
{
\begin{minipage}{\columnwidth}
\centering
\renewcommand*{\arraystretch}{1.5}
\tabcolsep=0.12cm
\begin{tabular}{|c| c|}
\hline %inserts horizontal line
Function Name & Equation/Description \\[0.5ex] 
\hline
Moment of Inertia & $\begin{array} {lcl} I & = & \sum\limits_{i=1}^n m_ir_i \end{array}$ \\[0.5ex]
\hline
Moment of Inertia on z axis & $\begin{array} {lcl} I_z & = & \sum\limits_{i=1}^n m_ir_{zi} \end{array}$ \\[0.5ex]
\hline
Sum of masses & $\begin{array} {lcl} M & = & \sum\limits_{i=1}^n m_i \end{array}$ \\[0.5ex]
\hline
Center of mass & $\begin{array} {lcl} CoM & = & \frac{I}{M} \end{array}$ \\[0.5ex]
\hline
Radius of Gyration & $\begin{array} {lcl} RG & = & \sqrt{\frac{I_z}{M}} \end{array}$ \\[0.5ex]
\hline
Dipole Moment& $\begin{array} {lcl} D & = & \sum\limits_{i=1}^n q_ir_i \end{array}$ \\[0.5ex]
\hline
Dipole Histogram & $\begin{array} {lcl} D_z & = & \sum\limits_{i=1}^n \frac{D}{z} \end{array}$ \\[0.5ex]
\hline
Electron Density & $\begin{array} {lcl} ED & = & \frac{\sum\limits_{i=1}^n (e_i-q_i)}{dz \cdot x \cdot y} \end{array}$ \\[0.5ex]
\hline
Heat Capacity & $\begin{array} {lcl} HC & = & \frac{3000 \cdot \sqrt{T} \cdot boltz}{2 \cdot \sqrt{T}-n \cdot df \cdot VarT} \end{array}$ \\[0.5ex]
\hline
Isothermal Compressibility & $\begin{array} {lcl} I & = & \frac{VarV}{V_{avg} \cdot boltz \cdot T \cdot PresFac} \end{array}$ \\[0.5ex]
\hline
Mean Square Displacement& $\begin{array} {lcl} msd & = & \langle (r_{t+\Delta_t}-r_t)^2 \rangle \end{array}$ \\[0.5ex]
\hline
Diffusion Constant & $\begin{array} {lcl} D_t & = & \frac{6 \cdot msd(t)}{t} \end{array}$ \\[0.5ex]
\hline Velocity Autocorrelation & $\begin{array} {lcl} V_{acor} & =&
\langle (V_{t+\Delta_t}\cdot V_t) \rangle \end{array}$ \\[0.5ex]
\hline
Force Autocorrelation & $\begin{array} {lcl} F_{acor} & = & \langle (F_{t+\Delta_t} \cdot F_t) \rangle \end{array}$ \\[0.5ex]
\hline
Density Function & Histogram of atom counts \\[0.5ex]
\hline
SDH & Histogram of all distances \\[0.5ex]
\hline
RDF & $\begin{array} {lcl} rdf(r) & = & \frac{SDH(r)}{4 \cdot \pi \cdot r^2 \cdot \sigma_r \cdot \rho} \end{array}$ \\[0.5ex]
\hline
\end{tabular}
\caption[Popular analytical queries in MS]{Popular analytical queries in MS }
\label{tb:queries}

@@

\section{One body functions}
\emph{\textbf{One body functions}} such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.

@@

The proposed design has a pretty straightforward structure which is going to be explained in this section. Shortly, the system starts with original input files (or data) and gets consumed in a sequential manner by simple disk IO framework into random access memory by chunks, then, chunk gets uploaded to GPU device and processed. At some point, the results that need to be shown to the user are transported from global memory of GPU (where they initially get accumulated) to the main memory, and then might get written as well to an output file. General data flow map is represented in Figure  \ref{fg:data-flow}.

@@

\section{Data input}
Since the size of the data in the problem is huge, organization of the data flow is crucial. As mentioned before, original data is retrieved from simulation software output files and in our case it is Gromacs Software's trajectory files\footnote{The file types, their structure, and their decompression has been well explained in the previous paper in section "Data orgranization in main memory"}. For a simulation, there can be several files of different format and usually they are also compressed without any loss of information. Tool for decompressing those files to a flat format for each frame and atom has been developed in the previous work\cite{mainPaper}.

@@

MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of queries. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

@@

Once the data is loaded into global memory of GPU, we have two kernels for two types of queries setup for two different streams. Besides it, we allocate some memory in GPU global memory for results, too, and keep it as an object with fields to easily pull it back when it is ready, in our case all the data sizes are static before the kernels are executed, which follows nature of GPU devices.

@@

As it has already been mentioned in previous sections, one body functions are relatively easy to implement, since they don't require iteration through data more than once with a write approach. Our implementation of one-body functions are relatively easy to understand which is what we tried to explain further.

@@

\subsection{SDH resolution performance impact}

>>>>

\titleformat{\chapter}[hang] 
{\centering\normalfont\bfseries}{\MakeUppercase{\chaptertitlename}\ \thechapter:}{0.5em}{\uppercase}

\titleformat{\section}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{0.5em}{}

\makeatletter

\newcommand\frontmatter{%
    \cleardoublepage
  %\@mainmatterfalse
  \pagenumbering{roman}}

\newcommand\mainmatter{%
    \cleardoublepage
 % \@mainmattertrue
  \pagenumbering{arabic}}

\newcommand\backmatter{%
  \if@openright
    \cleardoublepage
  \else
    \clearpage
  \fi
 % \@mainmatterfalse
   }

\makeatother

@@

\begin{document}
{
\frontmatter

\thispagestyle{empty}

@@

\addcontentsline{toc}{chapter}{ABSTRACT}

@@

\chapter*{Abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

@@

In this paper, we propose improved version of Push-based System for Molecular Simulation Data Analysis. Its major difference with the previous design is usage of GPU for the actual processing part of the data flow. Using the same scan-based I/O framework the data is pushed through the network of queries which are processed by GPU, and due to the nature of science simulation data, this gives a big advantage for processing it faster and easier (it will be explained more in later sections). In the old approach there were some custom data structures such as quad-tree for calculation of histograms to make the processing faster and those involved loss of data and some expectations from the data nature, too. In the new approach due to high performance of GPU processing and its nature, custom data structures were not even needed much, though it didn't bear any loss in precision and performance.

@@

\mainmatter
\linespread{1}\selectfont

\chapter{Introduction} \linespread{2}\selectfont \justify

\hspace{3em} In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

@@

The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms as mentioned and illustrated in previous work of the system \cite{mainPaper}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

@@

\section{Problem Statement}
\hspace{3em} Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Besides it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

@@

\hspace{3em} Reviewing the approach we are coming up with and what it was before helps us analyzing the weak parts and things that we can focus on to eliminate old bottlenecks.

@@

Finally, we've analyzed some main ideas, advantages, and disadvantages of our approach in this paper which is basically push-based system with optimization on data processing part by means of GPU computation (in other words, parallelism). Reading the data from files is a huge overhead, thus, speedup of push-based system in general is significant over pull-based system in our given case. 
It's been demonstrated in the previous paper with different amount of atoms, frames, and workloads. Since we have to load the data every time there is a different query versus push-based system loads a chunk of data once, it is quite noticeable how the framework contributes to efficiency of analysis of simulation data. Even though the memory loading and pushing steps are the key features of the proposed design, nevertheless, having general knowledge about the network of queries, in this paper, we will try to focus on optimizing the actual execution of them.

@@

\hspace{3em} In collaboration with Physics Department, the original motivation came from the need of improving analytic query executions performance over big data being generated by simulation software. The idea was to take input stream coming from simulation software (ideally live) and run network of queries very efficiently, because sometimes they might take days and more to compute. The proposed improved version of push-based system for molecular simulation data analysis, we believe, gives an opportunity for scientists to run their analysis fast having an ability to expand network of queries and customize computation functions to get the desired results. 
Our system primarily represents a system that works in several modules that were developed in C, CUDA, Python programming languages that, in combination, give us the desired result to perform efficient computations for the network of queries developed in the previous paper \cite{mainPaper} as well as to prove the concept of optimization of push-based system by means of parallelism with no data loss. We've developed the following modules:

@@

\hspace{3em} Push-based System for molecular simulation data analysis with its improved GPU features has couple of main focuses in terms of technical implementation and challenges related to those that have already been tried to be solved by other different approaches and software. Here is the review of other works.

@@

\hspace{3em} First of all, this system is a tool for scientific computation and in this case particularly for molecular simulation data analysis. PeriScope is a project and initiative by University of Michigan which is a system that will optimize data management for scientific explorations. Relational Database Management Systems were created to optimize business applications and it was revolutionary. Those couldn't be built without these systems with abstract languages like SQL, since it manipulates and automatically takes use of indexes, etc. PeriScope is a project that tries to come up with same approach against scientific data and computation with respectively larger scale.\cite{PeriScope} It is kind of difficult to compare our proposed system with PeriScope, but the original motivation is very similar: optimize scientific computation. PeriScope is a larger project directed towards more abstract solution like RDBMs that would work for any type of data. Another existing tool for scientific data and optimized processing is SciDB. It was created to make scientific data storage and querying better and faster. It is very similar to traditional RDBMS systems in terms of querying language and represent data storage in tables, but states to be faster than Postgres in 2 orders of magnitude. Some of the main features is distributed processing with atomicity and durability (the latter is quite difficult for large scale data), while the disadvantages are no-overwriting (only append mode), single value columns only (no arrays, etc), where records are like simple tuples with join function on querying multiple tables \cite{SciDB}. One of the major differences between Push-based Molecular simulation data analysis tool and the projects above is that our proposed design doesn't do anything with storing the data, because we are mainly concentrating it as a side plugin for processing and querying data in an optimized way, while the actual data generation and storage will be done with the actual simulation software.

\section{Stream processing}
\hspace{3em} Another major modern issue is live data streaming and live processing, which is also possible with our system. Basically, streaming function is a big challenge and need nowadays, because such networks as Twitter and Facebook are able to keep track of live trends and events, including people's reaction and feedback that have to be reflected real time for tremendous amount of users across the globe. Borealis Stream Processing Software is a tool being developed by MIT, Brown, and Brandeis Universities. It's primary goal is a distributed stream processing and based on Medusa and Aurora. Some of the major functions of the software are highly available distributed large streams processing live, dynamic and highly-scalable with possibility of changing queries live and dynamic vision of query results. \cite{borealis} There is also a similar need in scientific world, since being able to keep track of results in real time might be crucial to be able to make certain explorations.

@@

\hspace{3em} We are potentially dealing with data that could easily get up to terabytes and petabytes in sizes, and it will be fair to apply term of big data. Traditional software tools are not able to process Big Data apparently because of huge data sets, thus, special frameworks were developed to handle them. Nowadays there are bunch of frameworks that try to solve this problem. Some of these frameworks are Hadoop \cite{hadoop}, Hadoop with GPU extension \cite{hadoopgpu}, Spark \cite{spark}, Flink \cite{flink} (this is more of a big data stream processing framework) and other ones that are possibly just custom modifications and distributions of these. The mentioned tools help dealing with big data giving an ability to process them in batches on clusters using underlying HDFS (Highly Distributed File System), and normally go along with other utilities that help to apply MapReduce concept. The relationship between our system and this concept along with its implementations is in data sets sizes. Even though we are not using or introducing any non-traditional file systems, in certain scale and problems it could be required, too. In fact, the frameworks listed above could even be used for our problem on top of our system, because, as we discuss in section \ref{sc:conclusion}, it could be extended to a distributed system for analyzing independent frames in parallel.

\section{GPU and Big Data Processing}
\hspace{3em} GPU is becoming more popular being used by not only graphics processing, but machine learning, deep learning, search algorithms, sorting, and so on. Having an ability to optimize in tremendous speedups it is being involved in video classifications, natural language processing. Particularly about distributed processing and GPU applications along with it, there has been made some overview and research on what is the current state for common solutions in industry. "There were a lot of custom GPU-Hadoop frameworks after the launch of Mars project. These include C-MR, SteamMR, GPMR, Grex, Panda, Shredder and many others. However, most of these frameworks are no longer supported and were developed for some particular scientific projects. Therefore, a Monte Carlo simulation framework can be hardly applied for a bioinformatics project (say) based on the other algorithms. Moreover, processor technologies are evolving very rapidly. Many new revolutionary architectures have been developed for Sony PlayStation 4, Mali GPU by ARM, Adapteva Multicore Microprocessor etc. Both Mali GPU and Adapteva will be compatible with OpenCL. Xeon Phi co-processor works with OpenCL too, which was launched by INTEL. It has an x86-like architecture and is a 60 core co-processor that supports the PCI Express standard. It consumes just 300watts of energy and provides by giving a performance of 1 TFLOPS in double precision. Tianhe-2 which is the most powerful supercomputer till date implements this co-processor. Though, it is very difficult determine which Processor Computing framework architecture will more robust and high
performance." \cite{hadoopgpu}

@@

\chapter{Benchmark scientific queries}\label{sc:querynetwork}
\hspace{3em} Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries that were already introduced in the previous work of Push-Based System for Molecular Simulation Data Analysis \cite{mainPaper}.

@@

\vspace{3em}

\begin{table}[h]
\centering
\resizebox{\columnwidth}{!}
{
\begin{minipage}{\columnwidth}
\centering
\renewcommand*{\arraystretch}{1.5}
\tabcolsep=0.12cm
\begin{tabular}{|c| c|}
\hline %inserts horizontal line
Function Name & Equation\\[0.5ex] 
\hline
Moment of Inertia & $\begin{array} {lcl} I & = & \sum\limits_{i=1}^n m_ir_i \end{array}$ \\[0.5ex]
\hline
Sum of masses & $\begin{array} {lcl} M & = & \sum\limits_{i=1}^n m_i \end{array}$ \\[0.5ex]
\hline
Center of mass & $\begin{array} {lcl} CoM & = & \frac{I}{M} \end{array}$ \\[0.5ex]
\hline
Radius of Gyration & $\begin{array} {lcl} RG & = & \sqrt{\frac{I_z}{M}} \end{array}$ \\[0.5ex]
\hline
Dipole Moment& $\begin{array} {lcl} D & = & \sum\limits_{i=1}^n q_ir_i \end{array}$ \\[0.5ex]
\hline
Dipole Histogram & $\begin{array} {lcl} D_z & = & \sum\limits_{i=1}^n \frac{D}{z} \end{array}$ \\[0.5ex]
\hline
Density Function & Histogram of atom counts \\[0.5ex]
\hline
SDH & Histogram of all distances \\[0.5ex]
\hline
RDF & $\begin{array} {lcl} rdf(r) & = & \frac{SDH(r)}{4 \cdot \pi \cdot r^2 \cdot \sigma_r \cdot \rho} \end{array}$ \\[0.5ex]
\hline
\end{tabular}
\caption[Some of the popular analytical queries in MS]{Popular analytical queries in MS }
\label{tb:queries}

@@

\section{One body functions}
\hspace{3em} \emph{\textbf{One body functions}} such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.

@@

\hspace{3em} The proposed design has a pretty straightforward structure which is going to be explained in this section. Shortly, the system starts with original input files (or data) and gets consumed in a sequential manner by simple disk IO framework into random access memory by chunks, then, chunk gets uploaded to GPU device and processed. At some point, the results that need to be shown to the user are transported from global memory of GPU (where they initially get accumulated) to the main memory, and then might get written as well to an output file. General data flow map is represented in Figure  \ref{fg:data-flow}.

@@

\section{Data input}
\hspace{3em} Since the size of the data in the problem is huge, organization of the data flow is crucial. As mentioned before, original data is retrieved from simulation software output files and in our case it is Gromacs Software's trajectory files\footnote{The file types, their structure, and their decompression has been well explained in the previous paper in section "Data orgranization in main memory"}. For a simulation, there can be several files of different format and usually they are also compressed without any loss of information. Tool for decompressing those files to a flat format for each frame and atom has been developed in the previous work\cite{mainPaper}.

@@

\hspace{3em} MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of queries. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

@@

\hspace{3em} Once the data is loaded into global memory of GPU, we have two kernels for two types of queries setup for two different streams. Besides it, we allocate some memory in GPU global memory for results, too, and keep it as an object with fields to easily pull it back when it is ready, in our case all the data sizes are static before the kernels are executed, which follows nature of GPU devices.

@@

\hspace{3em} As it has already been mentioned in previous sections, one body functions are relatively easy to implement, since they don't require iteration through data more than once with a write approach. Our implementation of one-body functions are relatively easy to understand which is what we tried to explain further.

@@
